A global sensitivity analysis of model uncertainty in aeroelastic wind turbine models

### 2020-01-29 Papers/Torque_Paper/article.tex

\begin{abstract}
Sensitivity analysis of model parameters associated with the Blade element momentum models is studied. Sobol based 
\end{abstract}
\begin{keyword}
BEM, global sensitivity analysis, model uncertainty, Sobol indices
\end{keyword}
\end{frontmatter}

@@

The goal of this work is to perform a systematic assessment of the uncertainty in model parameters in BEM models. We approach this by performing a global sensitivity analysis based on the Sobol expansion approach, which decomposes the total variance of the quantity of interest (model output) into contributions from individual parameters and their combinations, similar to \cite{Echeverria2017,Murcia2018,Rinker2016}. We will employ the uncertainty quantification toolbox \texttt{UQLab} \cite{uqlab}, which computes the Sobol indices based on a sparse polynomial chaos expansion.

@@

\subsection{Geometric uncertainty}
We use Non-Uniform Rational Basis Splines (NURBS) \cite{rogers2000} to perturb the geometrical parameters of the turbine blade, such as the reference chord, twist or experimentally obtained drag-, lift-polar curves. The main advantage of using NURBS is that it provides great flexibility to approximate a large variety of curves with a limited number of control points which can be useful in reducing the number of uncertain parameters. %Further, the set of control points and knots can be directly manipulated to control the smoothness and curvature.
Use of NURBS has already been exploited in many aerodynamics applications, for instance, design optimization of wind-turbine blades, see for e.g. \cite{Bottasso2014,Ribeiro2012} and to parametrize chord and twist curves \cite{Echeverria2017}. In the following, we briefly outline the procedure used to parametrize curves using NURBS.
\subsubsection{NURBS based perturbation}
The value of NURBS curve $S(x)$ is computed using a weighted sum of $n$ basis functions (or B-splines):
\begin{equation}\label{NURB_curve}
S(x) = \sum_{i=1}^{n} c_i B_{i,p}(x),
\end{equation}
where $c_i$ is the weight of the $i$-th control point and $B_{i,p}(x)$ is the value of B-spline corresponding to the $i$-th control point at $x$ with the subscript $p$ denoting its polynomial degree. A sequence of non-decreasing parameters $X = \{x_1,x_2,...,x_m\}$ called the \textit{knot vector} determines the domain in which any control point is active with the number of elements in the knot vector is given by $m = n+p+1$. A B-spline $B_{i,p}(x)$ is non-zero only within a knot interval $[x_i,x_{i+1})$ for $i = 1, 2, ..., m-1$ and are zero elsewhere. Therefore, only a few nearby control points are active when computing the NURBS curve \eqref{NURB_curve}. For a curve of polynomial degree $p$, there are $p+1$ active control points at any location $x$, for e.g., for a linear curve ($p=1$) only two adjacent control points are active at any location. Also, note that the number of B-splines are equal to the number of elements in $X$. Another useful property of the B-splines is that they are recursive in polynomial degree thus we can derive quadratic B-splines using linear B-splines, cubic from quadratic B-splines and so on. Starting with B-spline of degree 0 defined as 
\begin{equation}\label{linearBspline}
B_{i,0}(x) :=
\begin{cases}
1\quad x_i\leq x < x_{i+1},\quad i = 1,2,...,m,\\
0\quad\text{elsewhere,}
\end{cases} 
\end{equation}
we can derive higher order B-splines in the knot interval $[x_{i+p},x_{i+p+1})$ using the recurrence relation \cite{deBoor}:
\begin{equation}\label{NURBS_recurrence}
B_{i,p}(x) := \frac{x - x_i}{x_{i+p} - x_i}B_{i,p-1}(x) + \frac{x_{i+p+1} - x}{x_{i+p+1} - x_{i+1}}B_{i+1,p-1}(x),\quad p\geq1.
\end{equation} 
These B-splines can be constructed efficiently using the de Boor's algorithm \cite{deBoor}. Note that to perform above computation, we need to pad the knot vector $X$ by repeating the first and last elements $p$ times each. In Fig. \ref{B-splines}, we show linear, quadratic and cubic splines for $x\in[0,1]$ [Also show the padding].

Next, we describe steps to generate perturbed samples of chord from a given reference chord, $S_{ref}(x)$, using NURBS based parametrization. The first step is to approximate the given reference chord using a NURBS curve with a fixed degree $p$ and a limited number of knots $X =\{x_j\}_{j=0}^n$. The knots can be uniformly spaced or chosen heuristically, more advanced approaches for knot selection can be found in, for e.g. \cite{LI200579,razdan1999knot}. Next we sample $S_{ref}(x)$ at $X$ and also compute $B_{i,p}(x_j), \text{ for }i, j = 1,2, ..., n$. Now we can formulate a linear system to solve for a set of control points $\mathbf{c}=\{c_i\}_{i=0}^n$ as:
\begin{equation}\label{nurbs_inversion}
\mathbf{B}\mathbf{c} = \mathbf{S}_{ref},
\end{equation}
where $\mathbf{S}_{ref}\in\mathbb{R}^{n}$ is a vector containing sampled values of $S_{ref}(x)$ and $\mathbf{B}\in \mathbb{R}^{n\times n}$ is a matrix with $j$-th row consisting of values of $n$ B-splines sampled at the location $x_j$. Once the control points are obtained, we can derive the approximate reference curve $S_n(x)$ using \eqref{NURB_curve}. In this work, we use uniform random variable to perturb the baseline control points $\mathbf{c}$.

There are two ways to obtain perturbed curves either by independently perturbing each of the control points (local perturbation) or using the same random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch1-Ch5 and twist control points Tw1-Tw6 control points, see Fig. \ref{perturbed_samples} (a) - (b), respectively. Each control point are perturbed using a uniform random number with bounds given by $\pm5\%$ of the magnitude of a control point. For lift and drag curves, all control points are perturbed using the same uniform random number resulting in samples of the lift and drag curves as shown in Fig. \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this. Also note that when using a uniform distribution with bounds given by the relative magnitude of control points, we tend to obtain small perturbations for control points with values close to zero.
\begin{figure}[h!]
\centering
\includegraphics[trim={0 0cm 0 0cm},clip, scale=.75]{basis_example.eps}
\caption{}
\label{perturbed_samples}
\end{figure}

\section{Global sensitivity analysis}\label{sec:GSA}
The objective of sensitivity analysis is to quantify the relative significance of individual inputs (or in combination) and how variations in input values affect the output of interest. In engineering, sensitivity analysis can be employed for several reasons: to determine the stability and robustness of a computational model for input parameters, for simplification of stochastic models by fixing the insensitive parameters, and to guide data acquisition campaigns and experimental design to refine the data on sensitive parameters. Sensitivity analysis techniques can be classified as local and global methods, see \cite{RSmith}. In a local sensitivity analysis, individual parameters are perturbed around their nominal values allowing for the description of output variability only in a small neighborhood of nominal input values. Although, local approaches are widely employed due to their ease of implementation and low computational cost, they are unable to quantify the global behavior of nonlinearly parametrized models such as aeroelastic models. Global sensitivity approaches, on the other hand, consider the entire range of input values to compute output sensitivities. Therefore, global sensitivity analysis is more suitable for aeroelastic models considered in this work.

\subsection{Sobol analysis}
We employ variance-based Sobol decomposition to perform global sensitivity analysis. This approach allows for quantification of the relative importance of the input parameters on a scale of $[0,1]$ known as \emph{Sobol indices}. In a Sobol analysis, we express the total variance of the output in term of contributions from individual parameters and their combinations. For the ease of exposition, let us consider a nonlinear model:
\begin{equation}\label{nonlinear_model}
Y = f({\mathbf{z}}),
where ${\mathbf{z}} = [z_1, z_2, ..., z_M]\in \mathcal{D}_{\mathbf{z}}\in \mathbb{R}^M$ is a vector of random inputs. For simplicity, we assume input parameters are uniformly distributed i.e. $z_i \sim \mathcal{U}(0,1)$ and the support of input set is $\mathcal{D}_{\mathbf{z}} = [0,1]^M$ where $M$ is the total number input parameters. The Sobol decomposition is based on the hierarchical representation \cite{RSmith}:
\begin{equation}\label{sobol_decomp}
f(z_1, z_2, ..., z_M) = f_0+\sum_{i=1}^M f_i(z_i) + \sum_{1\leq<i<j\leq M} f_{ij}(z_i,z_j) + ... + f_{1, 2, ..., M}(z_1, z_2, ..., z_M),
where the zeroth-order function $f_0$ is the mean response of $f$, the first-order univariate functions $f_i(z_i)$ quantify independent contributions due to each random input, second-order functions $f_{i,j}(z_i,z_j)$ represent the effect of interaction between $z_i$ and $z_j$ on the response $f$. Note that the \eqref{sobol_decomp} is only valid for independent input parameters. Higher-order terms are interpreted in a similar manner. Formally, the zeroth-, first- and second-order terms are defined as:

@@

where ${\mathcal{D}_{\mathbf{z}}}^{M-1} = [0,1]^{M-1}$ and the notation $\mathbf{z}_{\sim \{i,j\}}$ denotes the vector having all the components of $\mathbf{z}$ except in the set $\{i,j\}$. We only consider second order expansion of \eqref{sobol_decomp}, as for a number of applications, the second order expansion is usually sufficient and higher-order terms have negligible effect on the output response. In case of wind turbine, suppose $y$ represent the power output that depends on say 3 random inputs: windspeed $(z_1)$, wind standard deviation $(z_2)$ and RPM $(z_3)$, then $f_0$ represents the mean power output considering all the three random inputs, $f_1(z_1)$ represent the independent contribution of windspeed on the power and $f_{1,2}(z_1,z_2)$ quantify the interactions of wind speed and wind standard deviation on power. Further, the total variance of $f(\mathbf{z})$ is defined as:

@@

As the expansion in \eqref{sobol_decomp} is not unique, some orthogonality conditions must be imposed, see \cite{Rabitz1999,SOBOL2001271} which allows to express the total variance as

@@

The first- and second-order Sobol indices are computed as

@@

f(\mathbf{z}) \approx f^{K}(\mathbf{z}) = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}),

@@

where $\mathbf{k}\in \mathbb{N}_0^M$ is a $M$-dimensional multi-index with the magnitude $|\mathbf{k}| = k_1+k_2 + ... + k_M$. The total number of terms $K$ in \eqref{eq:PCE} can be based on a standard truncation scheme where we use all polynomials in $M$ random variable with order less than or equal to $p$. More advanced truncation schemes such as the maximum interaction or a hyperbolic scheme can be employed depending on the application \cite{BlatmanThesis}. The multivariate polynomials $\Psi_{\mathbf{k}}(\mathbf{z})$ is orthogonal with respect to the joint probability distribution of random vector $\mathbf{z}$. The multivariate distribution is computed using a tensor product of univariate polynomials $\psi_{k_i}^{(i)}(z_i)$

@@

The choice of the univariate orthogonal polynomial $\psi_{k_i}$ depends on the type of random variable, $z_i$. For example, for uniform random variable the Legendre family of polynomials is used, see \cite{Xiu2002} for details. Exploiting the orthogonality of $\Psi_{\mathbf{k}}(\mathbf{z})$, the coefficients can be computed using the projection approach:
\begin{equation}\label{eq:projection}
w_{\mathbf{k}} = \mathbb{E}[\Psi_{\mathbf{k}}(\mathbf{z})\cdot f^{K}(\mathbf{z})]
The above expectation can be computed using a quadrature based integration methods such as the Gaussian quadrature, sparse quadrature rules (for e.g., Smolyak's tensor grid), etc. Quadrature based integration methods suffer from curse of dimensionality where the number of integration points rises exponentially with an increase in the number of dimensions. In the current work where we have a fixed computational budget, methods that minimizes the approximation error for a given number of samples are more practical. Next, we briefly describe two such methods to compute PCE coefficients based on a fixed number of model evaluations.
\subsubsection{Ordinary Least Square}
Instead of using projection based method one can also use least squares based approaches to compute coefficients \cite{Berveiller2006}. The PCE expansion \eqref{eq:PCE} can be expressed as a vector-matrix product
\begin{equation}\label{eq:LSM}
Y = f^{K}(\mathbf{z}) +\varepsilon_K = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}) + \varepsilon_K = \mathbf{w}^T\boldsymbol{\Psi}(\mathbf{z}) +\varepsilon_K,
where $\mathbf{w} = \{w_0,w_1,...,w_K\}^T$ is the coefficient vector, $\boldsymbol{\Psi}(\mathbf{z}) = \{\Psi_0(\mathbf{z}), \Psi_1(\mathbf{z}), ..., \Psi_K(\mathbf{z})\}^T$ is the matrix containing multivariate polynomials in $\mathbf{z}$ and $\varepsilon_K$ is the approximation error. The least-square minimization problem can then be formulated as:
\begin{equation}\label{eq:PCE_OLS}
\mathbf{\hat{w}} = \text{arg min }\mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right]
The coefficient vector $\mathbf{\hat{w}}$ can be computed using the Ordinary Least Square (OLS) method. Given $N$ model evaluations $\mathbf{Y} = \{f(\mathbf{z}_1), f(\mathbf{z}_2), ..., f(\mathbf{z}_N)\}$, we can obtain the coefficients as:
\begin{equation}\label{eq:OLS}
\hat{\mathbf{w}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y},
where $\mathbf{X}\in \mathbb{R}^{N\times K}$ is the design matrix where $\mathbf{X}_{i,j} = \Psi_{j}(\mathbf{z}_i)$, $i = 1,2, ..., N$ and $j = 0,1,2,..., K$. Although the OLS method allows to compute the coefficients using a limited number of model evaluations one the main drawback of this approach is that it may result in a PCE model consisting of high-order interaction terms that may lead to overfitting. For many engineering problems, low-order interactions between random input variables are more important. Therefore, the least square form in \eqref{eq:PCE_OLS} can be regularized such that a the minimization results in a low order sparse PCE model.

\subsubsection{Least Angle Regression}
Least angle regression (LARS) is one of the several methods can be applied to compute sparse PCE models \cite{BlatmanThesis,BLATMAN2011}.

The LARS algorithm can be summarized in the following steps:

\subsection{Sensitivity analysis with discrete random variable}

\section{Results}\label{sec:results}

\subsection{NM80}
 and the turbine is the 2MW NM80 turbine from the DANAERO project \cite{Troldborg2013} with a blade radius of 38.8 m. The data for lift ($Cl$) and drag ($Cd$) polars are available at four locations along the blade radius at 11.87 m, 17.82 m, 28.97 m, and 35.53 m. The reference value of polars is obtained from the wind-tunnel experiment with 3D corrections. Random samples of chord, twist, lift- and drag-polars are obtained by perturbing the control points with a uniformly distributed random variable. For chord and twist, we independently perturb control points at different locations as shown in Fig. \ref{fig:samples} (a) - (b). For $Cl$ and $Cd$, all control points are perturbed using the same random number.

\subsubsection{Verification of GSA method}
Convergence study of UQLab with an increasing number of samples for OLS and LARS.
\begin{figure}[h!]
\centering
\includegraphics[trim={0cm 0cm 0 0cm},clip, scale=1.3]{nurbs_approx.eps}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.22]{convergence_chord1_4.eps}{a}
        \caption{$M=4$}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.22]{convergence_chord1_6.eps}
        \caption{$M=6$}
    \end{subfigure}
      ~ 
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.22]{convergence_chord1_8.eps}
        \caption{ $M=8$}
 \end{subfigure}
    \caption{ Convergence of $PCE\_OLS$ and $PCE\_LARS$ with increasing number of uncertain inputs $M$ and number of samples $N$}
\end{figure*}

\begin{figure}[h!]
\centering
\includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.4]{SA_Power_chord_twist_Cl_Cd.eps}
\caption{}
\label{perturbed_samples}
\end{figure}

>>>>

\begin{abstract}
A global sensitivity analysis of model parameters associated with the Blade Element Momentum (BEM) models is performed. The Sobol indices based on a sparse polynomial  expansion are used as a measure of global sensitivities. The sensitivity analysis workflow is developed using the uncertainty quantification toolbox \texttt{UQLab} that is integrated with the ECN's \texttt{Aero-Module} aeroelastic code. Sensitivity studies are performed on the NM80 wind turbine model from the DANAERO project.
\end{abstract}
\begin{keyword}
BEM, global sensitivity analysis, model uncertainty, Sobol indices
\end{keyword}
\end{frontmatter}

@@

The goal of this work is to perform a systematic assessment of the uncertainty in model parameters in BEM models. We approach this by performing a global sensitivity analysis based on the Sobol expansion approach, which decomposes the total variance of the quantity of interest (model output) into contributions from individual parameters and their combinations, similar to \cite{Echeverria2017,Murcia2018, Rinker2016a}. We will employ the uncertainty quantification toolbox \texttt{UQLab} \cite{uqlab}, which computes the Sobol indices based on a sparse polynomial chaos expansion.

@@

\subsection{Geometric uncertainty}
We use Non-Uniform Rational Basis Splines (NURBS) \cite{rogers2000} to perturb the geometrical parameters of the turbine blade, such as the reference chord, twist or experimentally obtained drag-, lift-polar curves. The main advantage of using NURBS is that it provides great flexibility to approximate a large variety of curves with a limited number of control points which can be useful in reducing the number of uncertain parameters. %Further, the set of control points and knots can be directly manipulated to control the smoothness and curvature.
Use of NURBS has already been exploited in many aerodynamics applications, for instance, design optimization of wind-turbine blades, see for e.g. \cite{Bottasso2014,Ribeiro2012} and to parametrize chord and twist curves \cite{Echeverria2017}. In the following, we briefly outline the procedure used to parametrize a curve using NURBS.
\subsubsection{NURBS based perturbation}
The value of NURBS curve $S(x)$ is computed using a weighted sum of $n$ basis functions (or B-splines):
\begin{equation}\label{NURB_curve}
S(x) = \sum_{i=1}^{n} c_i B_{i,p}(x),
\end{equation}
where $c_i$ is the weight of the $i$-th control point and $B_{i,p}(x)$ is the value of B-spline corresponding to the $i$-th control point at $x$ with the subscript $p$ denoting its polynomial degree. A sequence of non-decreasing parameters $X = \{x_1,x_2,...,x_m\}$ called the \textit{knot vector} determines the domain in which any control point is active. A control point $c_i$ is said to be active at a location when the corresponding B-spline $B_{i,p}(x)$ is non-zero at $x$. B-splines are recursive in polynomial degree thus we can derive quadratic B-splines using linear B-splines, cubic from quadratic B-splines and so on. Starting with B-spline of degree 0 defined as 
\begin{equation}\label{linearBspline}
B_{i,0}(x) :=
\begin{cases}
1\quad x_i\leq x < x_{i+1},\quad i = 1,2,...,m,\\
0\quad\text{elsewhere,}
\end{cases} 
\end{equation}
we can derive higher-order B-splines using the recurrence relation \cite{deBoor}:
\begin{equation}\label{NURBS_recurrence}
B_{i,p}(x) := \frac{x - x_i}{x_{i+p} - x_i}B_{i,p-1}(x) + \frac{x_{i+p+1} - x}{x_{i+p+1} - x_{i+1}}B_{i+1,p-1}(x),\quad p\geq1.
\end{equation}
Note that with increasing $p$ the interval in which a B-spline is non-zero also become larger. For a curve of polynomial degree $p$, there are $p+1$ active control points at any location $x$, for e.g., for a linear curve ($p=1$) only two adjacent control points are active at any location. Therefore, only a few nearby control points are active when computing the NURBS curve \eqref{NURB_curve}. These B-splines can be constructed efficiently using the de Boor's algorithm \cite{deBoor}. To compute high-order B-splines in \eqref{NURBS_recurrence}, we need to pad the knot vector $X$ by repeating the first and last elements $p$ times each. In Fig. \ref{basis_examples}, we show linear, quadratic and cubic splines for $x\in[0,1]$ with original knot vector $X = \{0,1/3,2/3,1\}$, padded accordingly with the curve order, for e.g., for linear B-splines, we use the knot vector $X = \{0,0,1/3,2/3,1,1\}$.
\begin{figure}[h!]
\centering
\includegraphics[trim={0 0cm 0 0cm},clip, scale=.75]{basis_example.eps}
\caption{Examples of B-splines with original knot vector $X = \{0,1/3,2/3,1\}$.}
\label{basis_examples}
\end{figure}

Next, we describe steps to generate perturbed samples of chord from a given reference chord, $S_{ref}(x)$, using NURBS based parametrization. The first step is to approximate the given reference chord using a NURBS curve with a fixed degree $p$ and a limited number of knots $X =\{x_j\}_{j=1}^n$. The knots can be uniformly spaced or chosen heuristically, more advanced approaches for knot selection can be found in \cite{LI2005791,razdan1999knot}. Next we sample $S_{ref}(x)$ at $X$ and also compute $B_{i,p}(x_j), \text{ for }i, j = 1,2, ..., n$. Now we can formulate a linear system to solve for a set of control points $\mathbf{c}=\{c_i\}_{i=1}^n$ as:
\begin{equation}\label{nurbs_inversion}
\mathbf{B}\mathbf{c} = \mathbf{S}_{ref},
\end{equation}
where $\mathbf{S}_{ref}\in\mathbb{R}^{n}$ is a vector containing sampled values of $S_{ref}(x)$ and $\mathbf{B}\in \mathbb{R}^{n\times n}$ is a matrix with $j$-th row consisting of values of $n$ B-splines sampled at location $x_j$. Once the control points are obtained, we can derive the approximate reference curve $S_n(x)$ using \eqref{NURB_curve}. In this work, we use uniform random variable to perturb the baseline control points $\mathbf{c}$. In Fig. \ref{nurbs_approx}, we show a parameterization of  chord curve using 9 control points and second-order B-splines.
\begin{figure}[h!]
\centering
\includegraphics[trim={0cm 0cm 0 0cm},clip, scale=1.3]{nurbs_approx.eps}
\caption{Parametrization of chord using 9 control points and second-order B-splines.}
\label{nurbs_approx}
\end{figure}

Finally, to obtain random samples of the chord curve, first we associate a probability distribution for each control point. And then we independently sample each control point and use them in \eqref{NURB_curve} to compute a random samples of chord curve.

\section{Global sensitivity analysis}\label{sec:GSA}
The objective of sensitivity analysis is to quantify the relative significance of individual inputs (or in combination) and how variations in input values affect the output of interest. In engineering, sensitivity analysis can be employed for several reasons: to determine the stability and robustness of a computational model for input parameters, for simplification of stochastic models by fixing the insensitive parameters, and to guide data acquisition campaigns and experimental design to refine the data on sensitive parameters. Sensitivity analysis techniques can be classified as local and global methods, see \cite{RSmith}. In a local sensitivity analysis, individual parameters are perturbed around their nominal values allowing for the description of output variability only in a small neighbourhood of nominal input values. Although, local approaches are widely employed due to their ease of implementation and low computational cost, they are unable to quantify the global behavior of nonlinearly parametrized models such as aeroelastic models. Global sensitivity approaches, on the other hand, consider the entire range of input values to compute output sensitivities. Therefore, a global sensitivity analysis is more suitable for aeroelastic models considered in this work.

\subsection{Sobol analysis}
We employ variance-based Sobol decomposition to perform global sensitivity analysis. This approach allows for quantification of the relative importance of the input parameters on a scale of $[0,1]$ known as \emph{Sobol indices}. In a Sobol analysis, we express the total variance of the output in term of contributions from individual parameters and their combinations. For the ease of exposition, let us consider a nonlinear model:
\begin{equation}\label{nonlinear_model}
Y = f({\mathbf{z}}),
\end{equation}
where ${\mathbf{z}} = [z_1, z_2, ..., z_M]\in \mathcal{D}_{\mathbf{z}}\in \mathbb{R}^M$ is a vector of random inputs. For simplicity, we assume input parameters are uniformly distributed i.e. $z_i \sim \mathcal{U}(0,1)$ and the support of input set is $\mathcal{D}_{\mathbf{z}} = [0,1]^M$ where $M$ is the total number input parameters. The Sobol decomposition is based on the hierarchical representation \cite{RSmith}:
\begin{equation}\label{sobol_decomp}
f(z_1, z_2, ..., z_M) = f_0+\sum_{i=1}^M f_i(z_i) + \sum_{1\leq i<j\leq M} f_{ij}(z_i,z_j) + ... + f_{1, 2, ..., M}(z_1, z_2, ..., z_M),
\end{equation}
where the zeroth-order function $f_0$ is the mean response of $f$, the first-order univariate functions $f_i(z_i)$ quantify independent contributions due to each random input, second-order functions $f_{i,j}(z_i,z_j)$ represent the effect of interaction between $z_i$ and $z_j$ on the response $f$.  Higher-order terms are interpreted in a similar manner. Note that the \eqref{sobol_decomp} is only valid for independent input parameters.

Formally, the zeroth-, first- and second-order terms are defined as:

@@

where ${\mathcal{D}_{\mathbf{z}}}^{M-1} = [0,1]^{M-1}$ and the notation $\mathbf{z}_{\sim \{i,j\}}$ denotes the vector having all the components of $\mathbf{z}$ except in the set $\{i,j\}$. We only consider second-order expansion of \eqref{sobol_decomp}, as for a number of applications, the second-order expansion is usually sufficient and higher-order terms have negligible effect on the output response. In the case of wind turbine, suppose $y$ represent the power output that depends on say 3 random inputs: windspeed $(z_1)$, wind standard deviation $(z_2)$ and RPM $(z_3)$, then $f_0$ represents the mean power output considering all the three random inputs, $f_1(z_1)$ represent the independent contribution of windspeed on the power and $f_{1,2}(z_1,z_2)$ quantify the interactions of wind speed and wind standard deviation on power. Further, the total variance of $f(\mathbf{z})$ is defined as:

@@

As the expansion in \eqref{sobol_decomp} is not unique, some orthogonality conditions must be imposed  \cite{Rabitz1999,SOBOL2001271} which allows to express the total variance as

@@

The first- and second-order Sobol indices are then computed as

@@

Y= f(\mathbf{z}) \approx f^{K}(\mathbf{z}) = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}),

@@

where $\mathbf{k}\in \mathbb{N}_0^M$ is a $M$-dimensional multi-index with the magnitude $|\mathbf{k}| = k_1+k_2 + ... + k_M$ and $\Psi_{\mathbf{k}}(\mathbf{z})$ multivariate polynomial is computed using a tensor product of univariate polynomials $\psi_{k_i}^{(i)}(z_i)$

@@

where $k_i$ is the order of univariate polynomial. The multivariate polynomials $\Psi_{\mathbf{k}}(\mathbf{z})$ is orthogonal with respect to the joint probability distribution of random vector $\mathbf{z}$. The choice of the univariate orthogonal polynomial $\psi_{k_i}$ depends on the type of random variable, $z_i$. For example, for uniform random variable the Legendre family of polynomials is used, see \cite{Xiu2002} for details.

The total number of terms $K$ in \eqref{eq:PCE} can be based on a standard truncation scheme where we use all polynomials in $M$ random variable with order less than or equal to $p$. More advanced truncation schemes such as the maximum interaction or a hyperbolic scheme can also be employed depending on the application, see \cite{BlatmanThesis}.

The PCE coefficient $w_{\mathbf{k}}$ corresponding to $\Psi_{\mathbf{k}}(\mathbf{z})$ can be computed using the projection approach:
\begin{equation}\label{eq:projection}
w_{\mathbf{k}} = \mathbb{E}[\Psi_{\mathbf{k}}(\mathbf{z})\cdot f^{K}(\mathbf{z})].
\end{equation} 
The above expectation can be computed using a quadrature based integration methods such as the Gaussian quadrature, sparse quadrature rules (e.g. Smolyak's tensor grid), etc. Quadrature based integration methods suffer from curse of dimensionality where the number of integration points rises exponentially with an increase in the number of dimensions. In the current work where we have a fixed computational budget, methods that minimizes the approximation error for a given number of samples are more practical. Next, we briefly describe two such methods to compute PCE coefficients based on a fixed number of model evaluations.
\subsubsection{Ordinary Least Square}
Instead of using projection based method one can also use least squares based approaches to compute coefficients \cite{Berveiller2006}. The PCE expansion \eqref{eq:PCE} can be expressed as a vector-matrix product
\begin{equation}\label{eq:LSM}
Y = f^{K}(\mathbf{z}) +\varepsilon_K = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}) + \varepsilon_K = \mathbf{w}^T\boldsymbol{\Psi}(\mathbf{z}) +\varepsilon_K,
\end{equation} 
where $\mathbf{w} = \{w_0,w_1,...,w_K\}^T$ is the coefficient vector, $\boldsymbol{\Psi}(\mathbf{z}) = \{\Psi_0(\mathbf{z}), \Psi_1(\mathbf{z}), ..., \Psi_K(\mathbf{z})\}^T$ is the matrix containing multivariate polynomials in $\mathbf{z}$ and $\varepsilon_K$ is the approximation error. The least-squares minimization problem can then be formulated as:
\begin{equation}\label{eq:PCE_OLS}
\mathbf{\overline{w}} = \text{arg min }\mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right]
 \end{equation}
The coefficient vector $\mathbf{\overline{w}}$ can be computed using the Ordinary Least Squares (OLS) method. Given $N$ model evaluations $\mathbf{Y} = \{f(\mathbf{z}_1), f(\mathbf{z}_2), ..., f(\mathbf{z}_N)\}$, we can obtain the coefficients as:
\begin{equation}\label{eq:OLS}
\overline{\mathbf{w}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y},
\end{equation}
where $\mathbf{X}\in \mathbb{R}^{N\times K}$ is the design matrix where $\mathbf{X}_{i,j} = \Psi_{j}(\mathbf{z}_i)$, $i = 1,2, ..., N$ and $j = 0,1,2,..., K$. Although the OLS method allows to compute the coefficients using a limited number of model evaluations, one of the main drawback of this approach is that it may result in a PCE model that consists of high-order interaction terms that may lead to overfitting. For many engineering problems, low-order interactions between the input variables are more important. Therefore, the least squares form in \eqref{eq:PCE_OLS} can be regularized such that the minimization results is a low-order sparse PCE model.

\subsubsection{Least Angle Regression}
Least angle regression (LARS) algorithm \cite{efron2004} is one of the several methods can be applied to compute sparse PCE models \cite{BlatmanThesis, BLATMAN2011}. The key idea is to find the best set of polynomials from a given candidate set that minimizes the regularized form of \eqref{eq:PCE_OLS}:
\begin{equation}\label{regularized_LS}
\mathbf{\overline{w}} = \text{arg min }\mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right] + \lambda ||\mathbf{w}||_1.
\end{equation}
The LARS algorithm in the context of PCE can be summarized as follows. We start with all PCE coefficients for the candidate polynomials set to zero, and find the polynomial, say $\Psi_1(\mathbf{w})$ most correlated with the model evaluations. This polynomial enters the active set. Next, we take the largest step in the direction of $\Psi_1(\mathbf{w})$ until some other polynomial $\Psi_2(\mathbf{w})$ from the candidate set, has as much correlation with the current residual. The second polynomial is now also added to the active set. In the next step, we move both the coefficients in the active set towards their least-square values until a third polynomial from candidate set shows as much correlation as the current PCE model based on two polynomials show with the model evaluations. The number of iteration is given by the minimum of $\{K,N\}$. After every iteration a posteriori error, for instance, Leave-One-Out (LOO) cross-validation error is computed. The active set with the smallest LOO is then chosen as the best sparse PCE
model.

One of the main advantage of the LARS algorithm is that it works well in cases when $K>>N$, i.e. the number of polynomials in the candidate set is much more larger than the number of model evaluations.

\subsection{Sensitivity analysis with discrete random variable}
\section{Results}\label{sec:results}

\subsection{NM80}
As a reference wind turbine, we consider the 2MW NM80 turbine from the DANAERO project \cite{Troldborg2013} with a blade radius of 38.8 m. The data for lift ($Cl$) and drag ($Cd$) polars are available at four locations along the blade radius at 11.87 m, 17.82 m, 28.97 m, and 35.53 m. The lift and drag variables at these sections are numbered Cl1 - Cl4 and Cd1 - Cd4 respectively. The reference value of polars is obtained from the wind-tunnel experiment with 3D corrections. Random samples of chord, twist, lift- and drag-polars are obtained by perturbing the control points with a uniformly distributed random variable.

There are two ways to obtain perturbed curves either by independently perturbing each of the control points (local perturbation) or using the same random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch3 - Ch7 and twist control points Tw2 - Tw7 control points, see Fig. \ref{perturbed_samples} (a) - (b), respectively. Each control point are perturbed using a uniform random number with bounds given by $\pm5\%$ of the magnitude of a control point. For lift and drag curves, all control points are perturbed using the same uniform random number resulting in samples of the lift and drag curves as shown in Fig. \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this purpose. Also note that when using a uniform distribution with bounds given by the relative magnitude of control points, we tend to obtain small perturbations for control points with values close to zero. 
\begin{figure}[h!]
\centering
\includegraphics[trim={0.5cm 0cm 0 0cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

\subsubsection{Verification of GSA method}
We perform a convergence study for OLS and LARS to determine which method converges fastest with increasing dimension. For this, we use the parametrized chord from Fig. \ref{perturbed_samples}(a) with 9 control points. In Fig. \ref{convergence_dim}, we show the convergence behavior of the two methods with increasing number of uncertain inputs, consisting of 4, 6 and 8 control points and for different number of model evaluations. We see that both methods scales well with increasing dimensions, however, $PCE\_LARS$ exhibits faster convergence than $PCE\_OLS$ in high dimensions.
